On the Synchronization Power of the Openflow Data Plane

### 2015-03-29 inband-sync.tex
first iteration

%\section*{todo before submission}

@@

In contrast, the transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}.
For instance, a transaction allows to modify switch configurations in an interactive
and computational fashion, as well as to define complex notions
of conflicts.
Our implementation is based on distributed identifiers and makes use of two
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.
Thus, our paper also provides the missing link for the read-modify-write object
postulated in~\cite{cpc}.

Finally, we also show that interesting synchronization mechanisms may be included
in-band even without the bundle feature, and as a case study, show
how to automatically \emph{compose} updates originating \emph{from different controllers}.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
In Section~\ref{sec:motivation}, we discuss the limitations of interpreting
the Openflow bundle as a transaction, and motivate our approach.
We then present our approach in detail in Section~\ref{sec:main}, and discuss use cases in Section~\ref{sec:apps}.
Section~\ref{sec:compo} describes how to perform in-band composition.
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}. Some technical details are postponed
to the Appendix.

@@

\liron{I think that maybe the rest of this section should be explained by the following subsections, providing in depth description of the OpenFlow features.}
\liron{Another direction is to have a simplified explenation as now and to refference spesific sections of the standard in side our paper.}
\subsection{The flow entry modification command (FlowMod)}
\lstset{% general command to set parameter(s)
	basicstyle=\small, % print whole listing small
	keywordstyle=\color{black}\bfseries, % underlined bold black keywords
	identifierstyle=, % nothing happens
	commentstyle=\color{mygrey}, % white comments
	stringstyle=\ttfamily, % typewriter type for strings
	showstringspaces=false, % no special string spaces
	language=C}
\begin{lstlisting}
struct ofp_flow_mod {
	struct ofp_header header;
	uint64_t cookie; /* Opaque controller-issued identifier. */
	uint64_t cookie_mask; /* Mask used to restrict the cookie bits
		that must match when the command is
		OFPFC_MODIFY* or OFPFC_DELETE*. A value
		of 0 indicates no restriction. */
	/* Flow actions. */
	uint8_t table_id; /* ID of the table to put the flow in.
		For OFPFC_DELETE_* commands, OFPTT_ALL
		can also be used to delete matching
		flows from all tables. */
	uint8_t command; /* One of OFPFC_*. */
	uint16_t idle_timeout; /* Idle time before discarding (seconds). */
	uint16_t hard_timeout; /* Max time before discarding (seconds). */
	uint16_t priority; /* Priority level of flow entry. */
	uint32_t buffer_id; /* Buffered packet to apply to, or
		OFP_NO_BUFFER.
		Not meaningful for OFPFC_DELETE*. */
	uint32_t out_port; /* For OFPFC_DELETE* commands, require
		matching entries to include this as an
		output port. A value of OFPP_ANY
		indicates no restriction. */
	uint32_t out_group; /* For OFPFC_DELETE* commands, require
		matching entries to include this as an
		output group. A value of OFPG_ANY
		indicates no restriction. */
	uint16_t flags; /* Bitmap of OFPFF_* flags. */
	uint16_t importance; /* Eviction precedence (optional). */
	struct ofp_match match; /* Fields to match. Variable size. */
	/* The variable size and padded match is always followed by instructions. */
	//struct ofp_instruction_header instructions[0];
		/* Instruction set - 0 or more. The length
		of the instruction set is inferred from
		the length field in the header. */
};
\end{lstlisting}

\subsubsection{The CHECK\_OVERLAP flag}
- creating and entry with overlapping match (same priority) and different instructions fails.
\subsection{Bundling}
- failed command aborts the bundle.
"When the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the xid of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle."
\subsubsection{Bundle control messages}

%/* Message structure for OFPT_BUNDLE_CONTROL. */
\begin{lstlisting}
struct ofp_bundle_ctrl_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t type; /* OFPBCT_*. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	/* Bundle Property list. */
	struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}
\subsubsection{Bundle add message}

%/* Message structure for OFPT_BUNDLE_ADD_MESSAGE.
\begin{lstlisting}
* Adding a message in a bundle is done with. */
struct ofp_bundle_add_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t pad; /* Align to 64 bits. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	struct ofp_header message; /* Message added to the bundle. */
	/* If there is one property or more, 'message' is followed by:
	* - Exactly (message.length + 7)/8*8 - (message.length) (between 0 and 7)
	* bytes of all-zero bytes */
	/* Bundle Property list. */
	//struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}

\textbf{Useful features for synchronization.} The Openflow standard (as of \textbf{TODO}) offers four interesting
features which we will exploit to implement the transactional memory abstraction:

\begin{enumerate}
\item \textbf{Bundling:}
A bundle can be seen as a ``mini-transaction'':
a bundles aggregates multiple control messages
into one \textsc{FlowMod} message, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
This bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.

\item \textbf{Check Overlap:}
The Openflow standard also supports the \emph{conditional} creation of
flow entries. A specific $\textsf{OFPFF\_CHECK\_OVERLAP}$ flag
set in the flow creation message ensures
that the flow entry is created if and only if it does not \emph{conflict}
with any other flow entry currently in the flow table.
This  conflict detection mechanism, combined with bundling,
allows us to implement specific types of \emph{transactions} that may
atomically install multiple rules if certain compatibility conditions
are met.

\item \textbf{Cookies:} Cookies allow to name and retrieve individual
flow entries and to use masks to delete modify/delete multiple entries. Our synchronization mechanism will implement controller-specific cookies.

\item \textbf{Groups:} Groups of action buckets can be defined in a special group table. Each group has a unique id and any attempt to create a group with existing id or to delete un existing group fails (and can abort a bundle in process).
\end{enumerate}

@@

@@

@@

\section{Implementing CAS with FlowMod commands}\label{sec:without-groups}
This section describes an implementation without group tables.

>>>>

\section*{todo}

stefan add old notes, bundle size complexity (how to make it minimal: store rule in one table, select it with a pointer / selector),
practical implementation (not even noviswitch has bundles)

@@

In contrast, the transactional interface presented in this paper supports very general atomic
transactions over the switch configuration space, similar to classic transactional memory~\cite{stm-st95,tm-book}.
For instance, a transaction allows to modify switch configurations in an interactive
and computational fashion, as well as to define complex notions
of conflicts.
Our implementation is based on distributed identifiers and makes use of two
new synchronization objects for managing these identifiers:
a \emph{claim-id object}, used for claiming an available identifier in a given
(bounded or unbounded) set, and a \emph{read-write register object}, used for storing the
currently used identifier.
The claim, unclaim, and check-claim operations can be implemented individually without the Openflow
bundle feature.

Our paper therefore also provides the missing link for the read-modify-write object
postulated in~\cite{cpc}.

%Finally, we also show that interesting synchronization mechanisms may be included
%in-band even without the bundle feature, and as a case study, show
%how to automatically \emph{compose} updates originating \emph{from different controllers}.

@@

\paragraph{Organization}
The rest of the paper is organized as follows.
In Section~\ref{sec:background}, we review the basics of the  Openflow
protocol and discuss its features which are relevant for our work.
In Section~\ref{sec:motivation}, we discuss the limitations of interpreting
the Openflow bundle as a transaction, and motivate our approach.
We then present our approach in detail in Section~\ref{sec:main}, and discuss use cases in Section~\ref{sec:apps}.
%Section~\ref{sec:compo} describes how to perform in-band composition.
We review related literature in Section~\ref{sec:relwork}, and conclude
in Section~\ref{sec:conclusion}. Some technical details are postponed
to the Appendix.

@@

\liron{I think that maybe the rest of this section should be explained by the following subsections, providing in depth description of the OpenFlow features.}
\liron{Another direction is to have a simplified explenation as now and to refference spesific sections of the standard in side our paper.}

\textbf{The FlowMod Command.}
\emph{FlowMod} is the main messages used in Openflow 
to update the switch configuration. 
In a nutshell, \emph{FlowMod} specifies the type of flow table modification,
a match and an action. 
The standard behavior is that the new rule simply replaces an existing one
with exactly the same match, and is installed
in addition to an existing one if the match only
partially overlaps. The latter can result
in inconsistencies, especially if the old and the new rule
have the same priority. \stefan{liron, is this correct?}
\emph{FlowMod} can be used together with a the CHECK\_OVERLAP flag
of \emph{FlowMod}, in order to detect conflicts: if the flow spaces of an old
and the new rule overlap, 
that is, if there is an entry with overlapping match part but different action, 
 and if the two rules have the same priority, then 
\emph{FlowMod} fails.

As we will see, in order to implement our synchronization objects,
the \emph{FlowMod} and CHECK\_OVERLAP features are sufficient.

\textbf{Bundling.}
The Openflow bundle can be seen as a ``mini-transaction'':
a bundles aggregates multiple control messages (add, remove, and modify rule)
into one, so that either all of the control messages are
processed, or none of them is. In particular, if, for some reason, the
configuration request contained in one of the bundled messages cannot
be processed, all the other requests  are rejected and an error
message is sent to the controller that issued them.
This bundled processing can be, when properly configured, done in a
\emph{packet-atomic} way, so that each packet is processed according
to the switch configuration either \emph{before} or \emph{after} the
request was processed.
If used together with the CHECK\_OVERLAP flag, the bundle command also returns a reason 
if the modification failed: the failed command in the bundle:
a CHECK\_OVERLAP flag can be defined for each command in the bundle.

Concretely, when the commit fails, the switch must generate an error message corresponding to the
message that could not be applied. The error message must use the xid of the offending message,
the error data field corresponding to that message and the error code corresponding to the failure. If
multiple messages are generating errors, the switch may return only the first error found or generate
multiple error messages for the same bundle.

We will make use of bundling only for policy updates. 
Bundling is \emph{not required} to implement transactions.

%\textbf{Cookies:} Cookies allow to name and retrieve individual
%flow entries and to use masks to delete modify/delete multiple entries. 
%Our synchronization mechanism will implement controller-specific cookies.
%
%\item \textbf{Groups:} Groups of action buckets can be defined in a special group table. Each group has a unique id and any attempt to create a group with existing id or %to delete un existing group fails (and can abort a bundle in process).
%\end{enumerate}

@@

what is the poewr of bundling flowmods:
- no cookies: delete entries now by specifying exact match part

- alterantive impelmentation without: group is bounded resources, can have own table etc. dont have many controlelrs anyway so not much is needed; also maybe other policies need groups

@@

\cleardoublepage

\onecolumn

@@

\section{The FlowMod Command}

\lstset{% general command to set parameter(s)
	basicstyle=\small, % print whole listing small
	keywordstyle=\color{black}\bfseries, % underlined bold black keywords
	identifierstyle=, % nothing happens
	commentstyle=\color{mygrey}, % white comments
	stringstyle=\ttfamily, % typewriter type for strings
	showstringspaces=false, % no special string spaces
	language=C}
\begin{lstlisting}
struct ofp_flow_mod {
	struct ofp_header header;
	uint64_t cookie; /* Opaque controller-issued identifier. */
	uint64_t cookie_mask; /* Mask used to restrict the cookie bits
		that must match when the command is
		OFPFC_MODIFY* or OFPFC_DELETE*. A value
		of 0 indicates no restriction. */
	/* Flow actions. */
	uint8_t table_id; /* ID of the table to put the flow in.
		For OFPFC_DELETE_* commands, OFPTT_ALL
		can also be used to delete matching
		flows from all tables. */
	uint8_t command; /* One of OFPFC_*. */
	uint16_t idle_timeout; /* Idle time before discarding (seconds). */
	uint16_t hard_timeout; /* Max time before discarding (seconds). */
	uint16_t priority; /* Priority level of flow entry. */
	uint32_t buffer_id; /* Buffered packet to apply to, or
		OFP_NO_BUFFER.
		Not meaningful for OFPFC_DELETE*. */
	uint32_t out_port; /* For OFPFC_DELETE* commands, require
		matching entries to include this as an
		output port. A value of OFPP_ANY
		indicates no restriction. */
	uint32_t out_group; /* For OFPFC_DELETE* commands, require
		matching entries to include this as an
		output group. A value of OFPG_ANY
		indicates no restriction. */
	uint16_t flags; /* Bitmap of OFPFF_* flags. */
	uint16_t importance; /* Eviction precedence (optional). */
	struct ofp_match match; /* Fields to match. Variable size. */
	/* The variable size and padded match is always followed by instructions. */
	//struct ofp_instruction_header instructions[0];
		/* Instruction set - 0 or more. The length
		of the instruction set is inferred from
		the length field in the header. */
};
\end{lstlisting}

\section{Bundle Control Message}

%/* Message structure for OFPT_BUNDLE_CONTROL. */
\begin{lstlisting}
struct ofp_bundle_ctrl_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t type; /* OFPBCT_*. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	/* Bundle Property list. */
	struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}

\begin{lstlisting}
* Adding a message in a bundle is done with. */
struct ofp_bundle_add_msg {
	struct ofp_header header;
	uint32_t bundle_id; /* Identify the bundle. */
	uint16_t pad; /* Align to 64 bits. */
	uint16_t flags; /* Bitmap of OFPBF_* flags. */
	struct ofp_header message; /* Message added to the bundle. */
	/* If there is one property or more, 'message' is followed by:
	* - Exactly (message.length + 7)/8*8 - (message.length) (between 0 and 7)
	* bytes of all-zero bytes */
	/* Bundle Property list. */
	//struct ofp_bundle_prop_header properties[0]; /* Zero or more properties. */
};
\end{lstlisting}

\begin{comment}
\section{Implementing CAS with FlowMod commands}\label{sec:without-groups}
This section describes an implementation without group tables.
